---
title: "Bryce2"
author: "Bryce Smith"
date: "3/2/2021"
output: word_document
---

```{r}
library(lubridate)
library(GGally)
df_mod1 <- read.csv('Seattle.csv')
df_mod1$date <- ymd(substr(date,1,nchar(date) - 7))
df_mod1$basement <- ifelse(df_mod1$sqft_basement == 0, 'No Basement','Basement')
head(df_mod1,3)
```

```{r fig1, fig.height = 5, fig.width=5}
quant.columns <- c(3, 4, 5, 6,7, 8, 11, 12, 13, 14)
ggpairs(df_mod1[quant.columns],title = "Correlogram of Quant. Columns")
```
#Price is right skewed. Placing a log transform on it to reduce the skewness.

```{r}
plot(density(log(df_mod1$price)))
```
#Helps with the skewness of price. Going to proceed.
```{r fig1, fig.height = 4, fig.width=5}
df_mod1$price <- log(df_mod1$price)
ggpairs(df_mod1[quant.columns],title = "Correlogram of Quant. Columns")
```
#Correlations increased between price and bedrooms, bathrooms, sqft_living, floors, grade, sqft_above, sqft_basement
#Correlation decreased between price and condition

#Fitting Quantitative Variables one at a time based on correlation.

```{r}
lm_1 <- lm(df_mod1$price ~ sqft_living, data = df_mod1)
summary(lm_1)
```

#sqft_living is significant and helpful

#Adding Grade Scores to the model
```{r}
lm_2 <- lm(df_mod1$price ~ sqft_living + grade, data = df_mod1)
summary(lm_2)
```
#Both grade and sqft_living, very significant, 

#Going to add bathrooms to the model because sqft_above will definitely be very correlated to sqft_living.
```{r}
lm_3 <- lm(df_mod1$price ~ sqft_living + grade + bathrooms, data = df_mod1)
summary(lm_3)
```
#R-squared increased but bathrooms for the added last t-test has a high p-value, therefore, should not be added to a model with sqft_living and grade in there.

#Going to take out bathrooms and add in the indicator variables, waterfront, and the basement.

```{r}
lm_4 <- lm(df_mod1$price ~ sqft_living + grade + waterfront, data = df_mod1)
summary(lm_4)
```
#Increase in R^2

#Trying view instead of waterfront
```{r}
lm_5 <- lm(df_mod1$price ~ sqft_living + grade + view, data = df_mod1)
summary(lm_5)

```

#Trying Basement with view
```{r}
lm_6 <- lm(df_mod1$price ~ sqft_living + grade + view + basement, data = df_mod1)
summary(lm_6)
```
#Significant predictors, R^2 0.6081 with both view and basement indicators. 

#Looking at the residual plots for the full model lm_6

```{r}
par(mfrow = c(1,3))
plot(lm_6,c(1,2,4))
```
#Doesn't look too bad. Residuals v. Fitted are equally dispersed on either side of zero. QQplot is fairly linear. Cook's distance shows 244 could be concerning.

#Lets look at the residuals plotting against the predictors
```{r}
par(mfrow = c(2,2))
plot(sqft_living ,lm_6$residuals, xlab = "sqft_living", ylab = "residuals")
plot(grade ,lm_6$residuals, xlab = "grade", ylab = "residuals")
plot(view ,lm_6$residuals, xlab = "view", ylab = "residuals")
```

#Sparsity on the right, going to try transformation


#First plotting responses vs. fitted responses

```{r}
plot(lm_6$fitted.values, df_mod1$price)
abline(a = 0, b = 1)
```

#Collinearity for lm_6
```{r}
library(car)
vif(lm_6)
```
#Collinearity VIF values look good.


#Trying log(sqft_living)
```{r}
lm_7 <- lm(df_mod1$price ~ log(sqft_living) + grade + view + basement, data = df_mod1)
summary(lm_7)
```
#Decrease in R-squared.




